---
title: "04: tidypollute, Meet, `tidycensus`"
author: "Dr. Nelson Roque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04: tidypollute, Meet, `tidycensus`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting Started with `tidypollute` and `tidycensus`

## Goals of this Notebook

This notebook demonstrates how to **download and stack** EPA AirData files for **2023-2024** focusing on **Ozone (44201)** data, and merge with `tidycensus` information.

------------------------------------------------------------------------

The simplest request retrieves EPA AirData for a specific **analyte** (pollutant) over a given time range (e.g., download daily ozone (44201) data for 1991-1992).

```{r, msg=F, warning=F}

library(tidypollute)
ozone <- get_epa_airdata(analyte = "44201", start_year = 2023, end_year = 2024, freq = "daily")

```


```{r, msg=F, warning=F}

library(tidycensus)
library(sf)
library(dplyr)

# Get Census tract demographics (e.g., median income, race)
census_data <- get_acs(
  geography = "tract",
  variables = c(median_income = "B19013_001"),
  state = "PA",  # Adjust for your state
  year = 2020,
  geometry = TRUE
)

# Convert EPA data to spatial format
epa_sf <- st_as_sf(ozone, coords = c("longitude", "latitude"), crs = 4326)

# Ensure Census data has the same CRS as epa_sf
census_data_t <- st_transform(census_data, st_crs(epa_sf))  # Transform to match EPA data

# Perform spatial join
epa_census <- st_join(epa_sf, census_data_t) %>%
  filter(variable == "median_income") %>%
  janitor::clean_names()
```

```{r}
library(ggplot2)

# does income predict pollution more in some parts of PA than others? ----
agg_data = epa_census %>% mutate(
    est_cut = cut(
      estimate,
      breaks = c(0, 20000, 40000, 60000, 80000, 100000, Inf),
      labels = c("Low", "Low-Mid", "Mid", "Mid-High", "High", "Very High")
    )
  ) %>%
  mutate(year = as.factor(lubridate::year(date_local))) %>%
  group_by(geoid, year, est_cut) %>%
  summarize(mean_pollution = mean(arithmetic_mean, na.rm = TRUE))

```

```{r}
  ggplot(agg_data, aes(x = year, y = mean_pollution, 
                       color = est_cut, group = geoid)) +
  geom_point() +
  geom_line()+
  theme_minimal() +
  labs(title = "Ozone Pollution Trends, by Income Bracket, in Pennsylvania",
       y = "Mean Ozone Levels (ppm)", x = "Year") +
  theme(legend.position = "none") +
  facet_grid(.~est_cut)

```

---

## Summary

This notebook provides a **simple workflow** for retrieving EPA AirData and merge with Census data using `tidypollute` and `tidycensus`.

You can:

✅ **Retrieve and filter dataset links (to zip files).**\
✅ **Download specific pollutant data** (e.g., Ozone, Wind).\
✅ **Stack and process the downloaded files.**\
✅ **Scales up to download all available EPA AirData.**

For more details, check out [`tidypollute` documentation](https://nelsonroque.github.io/tidypollute/index.html).
