---
title: "tidypollute: Making of the Hex Logo!"
author: "Dr. Nelson Roque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidypollute: Making of the Hex Logo!}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Getting Started with `tidypollute`

## Goals of this Notebook

```{r, message=F, warning=F}

# >>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>

# Load necessary libraries
library(hexSticker)
library(ggplot2)
library(showtext)
library(dplyr)
library(lubridate)
library(tidypollute)  # Assuming this is your package

# >>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>

## Load Google Fonts (for a handwritten, natural style)
font_add_google("IBM Plex Sans", "specialelite")
#font_add_google("Zilla Slab", "specialelite")
#font_add_google("Special Elite", "specialelite")
showtext_auto()  # Enable text rendering for the sticker

```

```{r, message=F, warning=F}
# Get EPA air pollution data ----
df_1991_co <- tidypollute::epa_airdata_links %>%
  filter(year == 1991, unit_of_analysis == "daily", analyte == "42101") %>%
  tidypollute::download_stack_epa_airdata(urls = ., download = TRUE, stack = TRUE, output_dir = "data/")

# Filter for Tampa, Florida -----
df_1991_co_filt <- df_1991_co %>%
  filter(state_name == "Florida", city_name == "Tampa") %>%
  mutate(dt_wday = wday(date_local, label = TRUE))
```

```{r}
# Count records per day -----
df_1991_co_filt %>% 
  count(date_local) %>%
  head(.) %>%
  knitr::kable(.) %>% 
  kableExtra::kable_classic(.)
```

```{r, message=F, warning=F}

# Create the ggplot visualization ------
p <- ggplot(df_1991_co_filt, aes(x = date_local, y = arithmetic_mean)) +
  geom_jitter(size = 0.2, alpha = 0.4, color="#C0C0C0") +  # Smoky gray for pollution particulates
  geom_smooth(color = "royalblue", alpha = 0.2, size = 0.9) +
  theme_minimal(base_family = "specialelite") +  # Use Google font
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p

```

```{r, message=F, warning=F}
# Generate the hex sticker ------
sticker(
  subplot = p,
  h_fill = "white",        # Dark gray smog background
  h_color = "royalblue",       # Industrial brown border
  package = "tidypollute",
  p_color = "royalblue",       # Muted yellow text (air quality warnings)
  p_family = "specialelite",
  p_size = 20,               # Adjust text size
  p_y = 1.25,                 # Adjust text height position
  s_x = 1, 
  s_y = 0.85,       # Scale and position the plot
  s_width = 1.3, 
  s_height = 0.95,
  filename = "../man/figures/logo.png"
)

```

## Summary

I always wanted to know to programmatically make hex logos for my R packages. This notebook shows how to make a hex logo for the `tidypollute` package. The logo is based on air pollution data from the EPA for Tampa, Florida in 1991, where the package author was born.

The logo is a visualization of the daily average CO levels in Tampa for 1991. The logo is a hex sticker that can be used for the `tidypollute` package.
